---
title: "Visualising GSEA results"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{gsea_visualisation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

`bixverse` gives you ranked gene set enrichment through `calc_fgsea()` and
`calc_blitzgsea()`. This vignette covers what to do with the results: running
enrichment curves, enrichment maps, and the calibration diagnostics that only
blitzGSEA has.

For the methods themselves, see the
[gene set enrichment vignette](https://gregorlueg.github.io/bixverse/articles/gse_methods.html)
in the parent package. This one assumes you already have results in hand.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(data.table)
library(ggplot2)
library(magrittr)
```

## The data

`fgsea` ships a mouse ranked list and a set of Reactome pathways. 12,000 genes,
586 pathways, and enough real signal to make the plots worth looking at.

```{r}
#| label: data
#| eval: !expr requireNamespace("fgsea", quietly = TRUE)
data(examplePathways, package = "fgsea")
data(exampleRanks, package = "fgsea")

stats <- sort(exampleRanks, decreasing = TRUE)

gsea_res <- calc_fgsea(
  stats = stats,
  pathways = examplePathways,
  gsea_params = params_gsea(min_size = 15L)
)

head(gsea_res[, .(pathway_name, es, nes, pvals, fdr)])
```

## Enrichment plots

`plot_gsea_enrichment()` draws the classical GSEA figure: the running
enrichment score across the ranked list, a rug of where the pathway genes sit,
and the ranked statistic underneath. It returns a **named list**, one ggplot per
pathway, so you can plot a handful in one call.

```{r}
#| label: enrichment-plain
#| fig-cap: "Running enrichment score for the top two hits."
top_pathways <- gsea_res[1:2, pathway_name]

plots <- plot_gsea_enrichment(
  stats = stats,
  pathways = examplePathways,
  pathways_of_interest = top_pathways
)

plots[[1]]
```

Hand it the results table as well and it annotates each panel with the NES and
FDR. Worth doing: a running score curve on its own tells you the shape of the
enrichment but nothing about whether it is significant.

```{r}
#| label: enrichment-annotated
#| fig-cap: "Same pathway, now with NES and FDR from the results table."
plots <- plot_gsea_enrichment(
  stats = stats,
  pathways = examplePathways,
  pathways_of_interest = top_pathways,
  gsea_results = gsea_res
)

plots[[1]]
```

**Key arguments:**

| Argument | Description |
| --- | --- |
| `stats` | Named numeric vector of the gene level statistic. Sort it descending. |
| `pathways` | The full named list you ran the enrichment on, not just the ones you are plotting. |
| `pathways_of_interest` | Which pathways to draw. Every one has to be in `names(pathways)`. |
| `gsea_results` | Optional. Needs `pathway_name`, `nes` and `fdr`, which is what `calc_fgsea()` and `calc_blitzgsea()` both return. |
| `tick_size` | Height of the rug ticks. |
| `text_size` | Size of the NES/FDR annotation. Only does anything when `gsea_results` is given. |

Note the size filter: the function drops pathways with fewer than three genes
present in `stats`, and they disappear from the returned list without comment.
If you get back fewer plots than you asked for, that is why.

**Extracting the plotting data**

Alternatively, you can pull the underlying data out and build your own plot.
`get_gsea_enrichment_data()` returns the curve, the tick positions and the key
points per pathway.

```{r}
#| label: enrichment-data
enrichment_data <- get_gsea_enrichment_data(
  stats = stats,
  pathways = examplePathways,
  pathways_of_interest = top_pathways,
  gsea_results = gsea_res
)

str(enrichment_data[[1]], max.level = 1)

head(enrichment_data[[1]]$curve_dt)
```

## Enrichment maps

Ranked enrichment on a library like Reactome gives you a lot of hits that are
mostly the same genes wearing different names. An enrichment map makes that
redundancy visible: gene sets become nodes, similarity between them becomes
edges, and Louvain clustering groups the ones that are telling you the same
thing.

`enrichment_map_gsea()` builds the `igraph`. It takes `calc_fgsea()` output
directly.

```{r}
#| label: enrichment-map-igraph
enrichment_map <- enrichment_map_gsea(
  res = gsea_res,
  threshold = 1e-4,
  pathways = examplePathways,
  min_sim = 0.2,
  resolution = 1.0
)

enrichment_map
```

Nodes are sized by `log10` of the gene set size and coloured by NES. The graph
carries its own layout, so the two renderers below agree on where things sit.

```{r}
#| label: enrichment-map-ggraph
#| fig-cap: "Enrichment map, adaptive labelling."
plot_enrichment_map_ggraph(enrichment_map, label_nodes = "adaptive")
```

Labelling is the part you will actually tune. `"adaptive"` scales the number of
labels to the community size, which keeps big clusters readable; `"all"` labels
everything and works up to maybe thirty nodes; an integer takes the top N by
node size; `NULL` drops labels entirely.

```{r}
#| label: enrichment-map-labels
#| fig-cap: "Top five nodes by size, everything else left bare."
plot_enrichment_map_ggraph(enrichment_map, label_nodes = 5L)
```

Long Reactome names need wrapping. `...` goes straight to
`wrap_and_truncate()`, so `width` and `max_lines` are set when you build the
graph, not when you plot it.

```{r}
#| label: enrichment-map-wrap
#| fig-cap: "Shorter labels via wrap_and_truncate()."
enrichment_map_short <- enrichment_map_gsea(
  res = gsea_res,
  threshold = 1e-4,
  pathways = examplePathways,
  width = 25L,
  max_lines = 2L
)

plot_enrichment_map_ggraph(enrichment_map_short, label_nodes = "all")
```

For anything beyond a handful of nodes the static version stops being useful and
you want to hover. `plot_enrichment_map_visnetwork()` takes the same graph and
gives you an interactive widget, same colours and same layout.

```{r}
#| label: enrichment-map-visnetwork
plot_enrichment_map_visnetwork(enrichment_map)
```

## blitzGSEA

blitzGSEA calibrates a gamma null once against the signature and then reads
every pathway's p-value off the fitted tail. Cost per pathway collapses, which
is what makes it worth using on large libraries. The results table has the same
`pathway_name`, `nes` and `fdr` columns as `calc_fgsea()`, so everything above
works on it unchanged.

```{r}
#| label: blitz-run
#| eval: !expr requireNamespace("fgsea", quietly = TRUE)
blitz_params <- params_blitzgsea(min_size = 15L)

null_model <- blitzgsea_calibrate(
  stats = stats,
  blitz_params = blitz_params
)

blitz_res <- calc_blitzgsea(
  stats = stats,
  pathways = examplePathways,
  blitz_params = blitz_params,
  null_model = null_model
)

head(blitz_res[, .(pathway_name, es, nes, pvals, fdr, size)])
```

### Centre your stats before drawing the curve

One trap. `params_blitzgsea()` defaults to `centre = TRUE`, and the enrichment
score is not invariant to an offset: the walk weights each gene by `|stat|^p`,
so shifting the statistic changes the score. `plot_gsea_enrichment()` draws the
uncentred walk. Hand it the raw ranks next to `calc_blitzgsea()` output and the
curve peaks somewhere other than the `es` in the table.

Subtract the mean and they line up exactly:

```{r}
#| label: blitz-centring
top_blitz <- blitz_res[1, pathway_name]

peak_of <- function(s) {
  d <- get_gsea_enrichment_data(
    stats = s,
    pathways = examplePathways,
    pathways_of_interest = top_blitz
  )
  kp <- d[[1]]$key_points
  kp[which.max(abs(kp[c("pos_es", "neg_es")]))]
}

c(
  reported = blitz_res[1, es],
  raw = peak_of(stats),
  centred = peak_of(stats - mean(stats))
)
```

So centre first:

```{r}
#| label: blitz-enrichment
#| fig-cap: "blitzGSEA hit, drawn on the centred signature."
plot_gsea_enrichment(
  stats = stats - mean(stats),
  pathways = examplePathways,
  pathways_of_interest = top_blitz,
  gsea_results = blitz_res
)[[1]]
```

Running `calc_fgsea()` instead? Then leave the stats alone. Only blitzGSEA
centres.

### Is the calibration any good?

Everything blitzGSEA reports rests on the gamma fits describing the permutation
scores. `plot_blitzgsea_null()` shows the smoothed parameters across the anchor
grid, one line per tail, with the Kolmogorov-Smirnov goodness-of-fit p-values in
the subtitle.

```{r}
#| label: blitz-null
#| fig-cap: "Smoothed gamma parameters across the anchor grid."
#| fig-height: 7
plot_blitzgsea_null(null_model)
```

What you want is smooth curves and KS p-values comfortably above 0.05. Sharp
kinks or a fit that sags at one end of the size range mean the tails are
misbehaving for gene sets of that size, and the p-values there are optimistic.
More permutations is the usual fix. `calc_blitzgsea()` warns on its own when the
mean KS p-value drops below 0.05, but the warning is one number for the whole
grid and this plot tells you *where* it went wrong.

The calibration depends only on the signature, never on the library. Score five
libraries against one ranking and you pay for this once, which is why
`blitzgsea_calibrate()` hands the null back as a plain list you can `saveRDS()`.

### Where the p-value comes from

`plot_blitzgsea_es_null()` puts the observed score on the fitted null at that
pathway's size, with the tail beyond it shaded. That shaded area, doubled, is
the p-value in the results table.

```{r}
#| label: blitz-es-null
#| fig-cap: "Observed enrichment score against the fitted null."
es_plots <- plot_blitzgsea_es_null(
  null_model = null_model,
  res = blitz_res,
  pathways_of_interest = blitz_res[1:2, pathway_name]
)

es_plots[[1]]
```

The gamma parameters are interpolated log-linearly from the anchor grid to the
pathway's exact size, matching what the Rust does when it scores. This is the
fitted density, not a histogram: the permutation draws never leave Rust, so
there is no empirical null to overlay on it.

A non-significant pathway looks like what you would expect, the score sitting
in the body of the distribution:

```{r}
#| label: blitz-es-null-boring
#| fig-cap: "A pathway that is not going anywhere."
boring <- blitz_res[fdr > 0.5][1, pathway_name]

plot_blitzgsea_es_null(
  null_model = null_model,
  res = blitz_res,
  pathways_of_interest = boring
)[[1]]
```

## Related

Over-representation results (hypergeometric tests) get their own plotting
helpers, covered in the
[ORA vignette](https://gregorlueg.github.io/bixverse.plots/articles/ora_visualisation.html).
Both vignettes use the same dataset, so the enrichment maps there are directly
comparable to the ones here.
