---
title: "Visualising over-representation results"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{ora_visualisation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Over-representation analysis takes a discrete gene list and asks which gene sets
it overlaps more than chance would allow. `bixverse` does the hypergeometric
test through `gse_hypergeometric()` and `gse_hypergeometric_list()`. This
vignette covers the plotting side: dot plots and enrichment maps.

For the test itself and the GO-aware variants, see the
[gene set enrichment vignette](https://gregorlueg.github.io/bixverse/articles/gse_methods.html)
in the parent package.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(data.table)
library(ggplot2)
library(magrittr)
```

## The data

Same mouse ranking and Reactome pathways as the
[GSEA vignette](https://gregorlueg.github.io/bixverse.plots/articles/gsea_visualisation.html),
so the enrichment maps in the two are directly comparable. ORA needs a discrete
list, so we cut the ranking at the top 500 genes and use the full measured set
as the universe.

Picking the universe properly matters more than people give it credit for. All
12,000 measured genes is the right call here; the union of the library would
quietly assume every gene in Reactome was assayed.

```{r}
#| label: data
#| eval: !expr requireNamespace("fgsea", quietly = TRUE)
data(examplePathways, package = "fgsea")
data(exampleRanks, package = "fgsea")

stats <- sort(exampleRanks, decreasing = TRUE)
gene_universe <- names(stats)

# the numeric Reactome prefix carries nothing and the underscores make for
# ugly labels. Clean the names once, up front
pathways <- examplePathways
names(pathways) <- gsub("_", " ", gsub("^\\d+_", "", names(pathways)))

target_genes <- head(gene_universe, 500)

ora_res <- gse_hypergeometric(
  target_genes = target_genes,
  gene_set_list = pathways,
  gene_universe = gene_universe,
  threshold = 0.05
)

head(ora_res)
```

## Dot plots

`plot_gse_dotplot()` takes the results table straight from
`gse_hypergeometric()`.

```{r}
#| label: dotplot
#| fig-cap: "Over-represented Reactome pathways in the top 500 genes."
#| fig-height: 7
plot_gse_dotplot(ora_res)
```

Four things are encoded and only two of them are obvious, so: the x axis is the
gene ratio, `hits / target_set_lengths`, i.e. the fraction of your input list
that landed in the set. Pathways are ordered by that same ratio. Fill is FDR and
point size is the size of the gene set.

That last one is the pairing to read carefully. A big point far to the right is
a large gene set swallowing a large chunk of your list, which is usually less
interesting than a small point at the same x position.

**Key arguments:**

| Argument | Description |
| --- | --- |
| `res` | Results table. Needs `hits`, `target_set_lengths`, `gene_set_name`, `gene_set_lengths` and `fdr`. |
| `size_range` | Point size range, smallest to largest gene set. |
| `viridis_option` | Which viridis map to use for the fill, `"A"` through `"H"`. |
| `direction` | `1` or `-1`, flips the fill scale. |
| `max_terms` | Keep only the N most significant sets. Defaults to `NULL`, everything. |
| `.verbose` | Silences the message about splitting when several target sets are present. |
| `...` | Forwarded to `wrap_and_truncate()`, which shortens the labels. `width` and `max_lines` are the ones you want. |

### Several target sets at once

`gse_hypergeometric_list()` tests many lists in one go and adds a
`target_set_name` column. `plot_gse_dotplot()` picks that up and returns a named
list of plots rather than a single one.

ORA depends on where you cut the ranking, and it is worth seeing how much. Same
statistic, three thresholds:

```{r}
#| label: dotplot-list
target_list <- list(
  top_250 = head(gene_universe, 250),
  top_500 = head(gene_universe, 500),
  top_1000 = head(gene_universe, 1000)
)

ora_res_ls <- gse_hypergeometric_list(
  target_genes_list = target_list,
  gene_set_list = pathways,
  gene_universe = gene_universe,
  threshold = 0.05
)

ora_res_ls[, .N, by = target_set_name]
```

14, 24 and 54 significant pathways off one ranking, with 13 common to all three.
The cell cycle story survives every cut, which is reassuring, but the long tail
is entirely an artefact of where you drew the line. This is the reason to reach
for GSEA when you have a full ranking.

54 gene sets do not fit on a dot plot, and Reactome names are long enough to
swallow the panel whole. `max_terms` keeps the most significant N, and anything
in `...` goes to `wrap_and_truncate()` for the labels.

```{r}
#| label: dotplot-list-plots
#| fig-cap: "Top 250 genes, capped at the 20 most significant sets."
#| fig-height: 6
plots <- plot_gse_dotplot(
  ora_res_ls,
  max_terms = 20L,
  width = 40L,
  .verbose = FALSE
)

plots$top_250
```

```{r}
#| label: dotplot-list-plots-2
#| fig-cap: "Same treatment on the top 1000, where 54 sets came back."
#| fig-height: 6
plots$top_1000
```

## Enrichment maps

Reactome is a hierarchy, so significant hits arrive in correlated clumps. An
enrichment map makes that structure explicit: nodes are gene sets, edges are the
overlap between them, and Louvain clustering pulls out the groups.

`enrichment_map_oae()` builds the `igraph` from the ORA results. Nodes are
coloured by `-log10(FDR)` here, where the GSEA version colours by NES.

```{r}
#| label: map-igraph
enrichment_map <- enrichment_map_oae(
  res = ora_res,
  threshold = 0.05,
  pathways = pathways,
  min_sim = 0.2,
  resolution = 1.0
)

enrichment_map
```

```{r}
#| label: map-ggraph
#| fig-cap: "Enrichment map of the over-represented pathways."
plot_enrichment_map_ggraph(enrichment_map, label_nodes = "adaptive")
```

Three arguments change the picture, and they are all set when you build the
graph rather than when you plot it.

`min_sim` is the edge threshold. Raise it and weak overlaps vanish, which splits
the graph into more, tighter communities. On this data 0.1 gives 89 edges across
6 communities and 0.4 gives 32 edges across 11.

```{r}
#| label: map-min-sim
#| fig-cap: "min_sim at 0.4. Sparser, and the clusters fragment."
enrichment_map_strict <- enrichment_map_oae(
  res = ora_res,
  threshold = 0.05,
  pathways = pathways,
  min_sim = 0.4
)

plot_enrichment_map_ggraph(enrichment_map_strict, label_nodes = "adaptive")
```

`resolution` is the Louvain knob: higher gives more, smaller communities. It
only changes the colouring and how many labels adaptive mode shows, never the
edges.

`overlap_coefficient` swaps Jaccard for the overlap coefficient. Worth flipping
whenever your library mixes very different set sizes. Jaccard punishes a small
set nested inside a large one, the overlap coefficient scores it as 1, which is
usually what you meant for a hierarchy like Reactome.

```{r}
#| label: map-overlap
#| fig-cap: "Overlap coefficient instead of Jaccard."
enrichment_map_overlap <- enrichment_map_oae(
  res = ora_res,
  threshold = 0.05,
  pathways = pathways,
  overlap_coefficient = TRUE,
  min_sim = 0.5
)

plot_enrichment_map_ggraph(enrichment_map_overlap, label_nodes = "adaptive")
```

The interactive version takes the same graph, keeps the layout, and lets you
hover for the full pathway name rather than the truncated label.

```{r}
#| label: map-visnetwork
plot_enrichment_map_visnetwork(enrichment_map)
```

## Saving

`save_plot()` writes a single plot, `save_plot_ls()` writes a named list, which
is exactly what `plot_gse_dotplot()` hands back for a multi-target run. File
names come from the list names. Both take a `params_plots()` list.

```{r}
#| label: saving
#| eval: false
plot_params <- params_plots(
  width = 8,
  height = 6,
  file_type = ".png",
  res = 450L
)

save_plot(
  plot = plot_gse_dotplot(ora_res),
  path = file.path(tempdir(), "ora_dotplot"),
  plot_params = plot_params
)

# one file per target set
save_plot_ls(
  plot_ls = plots,
  path = file.path(tempdir(), "ora_dotplots"),
  plot_params = plot_params
)
```

## Related

Working from a full ranking rather than a cut list? The
[GSEA vignette](https://gregorlueg.github.io/bixverse.plots/articles/gsea_visualisation.html)
covers running enrichment curves, the GSEA flavour of the enrichment map, and
the blitzGSEA calibration diagnostics.
